Unsupervised Characterization of Water Composition with UAV-based Hyperspectral Imaging and Generative Topographic Mapping

### 2024-06-24 paper/paper.tex

In this paper, we explore the application of the GTM to UAV-acquired HSI for the characterization water quality. Using data collected at a pond in Montague, North Texas, we first train a GTM using water-only pixels to produce a low-dimensional representation of the collected HSI. We use this mapping to explore the highly detailed small-scale variability within the pond and discuss how this can be used to guide reference data collection. Next, we demonstrate how the GTM can be utilized to identify relevant spectral endmembers from a combined dataset including land pixels, algae, water, and a simulated contaminant plume using a rhodamine tracer dye. Once identified, these endmembers can be used to rapidly map the abundance of spectral features within the pond. We demonstrate this capability by using a GTM to map the abundance of filamentous blue-green algae near the shore as well as the dispersion of a rhodamine dye plume.

@@

In this study, we explore the use of the GTM as a tool for dimensionality reduction and unsupervised endmember extraction of hyperspectral imagery. To this end, a dataset of HSI was collected at a pond in Montague, North Texas (shown in panel d of Fig~\ref{fig:sample-spectra}) on 23 November 2020 using a UAV mounted hyperspectral imager configured as described in \cite{robot-team-1, robot-team-2}. In this section we first provide a detailed overview of the GTM algorithm. Next we describe the UAV platform used for HSI collection and the various steps used to process captured HSI. \textcolor{red}{Finally, we describe the two case studies presented in this paper for the utilization of the GTM as a HSI visualization tool and as an endmember extraction technique.}

@@

Points $\xi$ in this latent space are mapped to the embedded manifold by a non-linear function $\psi$. However, since real data is rarely noise-free, this embedded manifold will not be perfectly thin. To account for this, the points $\xi$ are described in the data space by a radially symmetric Gaussian distribution, $\mathcal{N}(\psi(\xi),\, \beta^{-1})$, with mean $\psi(\xi)$ and precision $\beta$. For a dataset $\left\{ \mathbf{x}_n\right\}\limits_{n=1}^N$ consisting of $N$-many records, this choice yields a log likelihood function of the form
\begin{equation}\label{eq:llh}
    \mathcal{L}(W, \beta) = \sum_n^N \ln \left(\dfrac{1}{K}\sum_k^K p(\mathbf{x}_n \mid \xi_k, W, \beta) \right)
\end{equation}
which can be maximized to obtain optimal $W$ and $\beta^{-1}$.

@@

Slices $R_{[:,n]}$ of the matrix $R$ define the responsibility of each latent node $\xi_k$ for the $n$th data record $\mathbf{x}_n$. Therefore the final responsibility matrix after GTM training can be used to represent each record in the latent space by the mean:
\begin{equation}
    \hat{\xi}_n = \sum_{k}^K R_{kn}\xi_k.
\end{equation}

@@

To account for the variability of incident light, raw HSI are converted to units of reflectance using the downwelling irradiance spectrum simultaneously captured with each HSI. With the hyperspectral imager oriented to nadir, the reflectance is given by
\begin{equation}\label{eq:reflectance}
    \rho(\lambda) = \pi L(\lambda)/E_d(\lambda)
\end{equation}
where $L$ is the spectral radiance, $E_d$ is the downwelling irradiance, and a factor of $\pi$ steradians results from assuming Lambertian (diffuse) upwelling radiance \cite{ruddick2019review}. HSI collection was performed near solar noon to maximize the amount of incident sunlight illuminating the water. For the site in North Texas, this corresponded to an average solar zenith angle of $54.9^\circ$ for 23 November 2020 resulting in HSI with negligible sunglint effects.

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=15.5cm]{paper/figures/methods/annotated-drone.pdf}
\end{adjustwidth}
\caption{The UAV platform: (\textbf{a}) The Resonon Pika XC2 hyperspectral imager. (\text{b}) The Freefly Alta-X with the attached hyperspectral imager, processing computer, and downwelling irradiance spectrometer. \label{fig:drone}}
\end{figure}

@@

Next, we consider a dataset of combined HSI pixels including both water and land. To simulate the dispersion of a potential contaminant source, a Rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume. From these HSI a collection of $>145,000$ pixels were sampled for model training. Exemplar spectra from water, ground, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}. Using these spectra, we explore the trained GTM and use it to extract spectral signatures corresponding to these endmember categories. Values for the hyperparameters $m$, $s$, and $\alpha$, are determined by training multiple GTM models and selecting values which minimize the Bayesian Information Criterion (BIC) given by 
\begin{equation}
    \text{BIC} = P\ln(N) - 2\mathcal{L}
\end{equation}
where $P$ is the total number of model parameters, $N$ is the number of records in the dataset, and $\mathcal{L}$ is the log likelihood defined in Eq~\ref{eq:llh}.

The normalized spectral similarity score (NS3) introduced by Nidamanuri and Zbell combines the root-mean-square (RMS) difference together with the spectral angle to provide a spectral distance function\cite{nidamanuri2010normalized}. For two spectra $\rho_1(\lambda)$ and $\rho_2(\lambda)$ it is defined by
\begin{equation}
    \text{NS3}(\rho_1, \rho_2) &= \sqrt{RMS(\rho_1, \rho_2)^2 + (1-\cos\theta)^2}
\end{equation}
where
\begin{align}
    \text{RMS}(\rho_1, \rho_2) &= \sqrt{\frac{1}{N-1}\sum_\lambda \left(\rho_{1}(\lambda) - \rho_2(\lambda) \right)^2} \\
    \cos\theta &= \frac{\langle \rho_1 , \rho_2 \rangle}{\lVert \rho_1\rVert \lVert \rho_2 \rVert}.
\end{align}
We use the NS3 together with identified spectral endmembers to map the abundance of algae near the shore as well as the evolution of the Rhodamine dye plume.

@@

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=18.0cm]{paper/figures/results/gtm-water.png}
\end{adjustwidth}
\caption{Classification map for GTM trained solely on water pixels (no land and no rhodamine plume). \textbf{(a)} GTM applied to all water pixels colored by their associated position in the latent space. \textbf{(b)} Representation of the points from the training set in the latent space. Color is assigned to each point by mapping $\xi_1$ to red and $\xi_2$ to blue. \textbf{(c)} GTM spectral signatures, $\psi(\xi_k)$, corresponding to nodes from the four corners and center of the GTM latent space. \label{fig:gtm-water}}
\end{figure}

@@

To visualize the distribution of HSI learned by the GTM, we can associate a color with each dimension of the latent space. In Fig~\ref{fig:gtm-water} we have used the red channel to represent $\xi_1$ and the blue channel to represent $\xi_2$. Applying the trained GTM to compute mean node responsibilities for all water-pixels in collected HSI allows us to illustrate the spatial distribution of the spectra on a map as shown in panel (a). Here we observe a clear distinction between water near the shore and water in the middle of the pond. Additionally, the eastern alcove of the pond is significantly more blue than the rest of the water reflecting the limited flow through this region. The close proximity of highly dissimilar GTM classes illustrated by sharp color gradients in the map captures the small-scale spatial variability typical of inland water bodies like this pond.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\columnwidth]{paper/figures/results/square-ndvi.pdf}
\caption{GTM latent space visualization: Each sample spectrum from the dataset of combined HSI covering shore, water, and the rhodamine dye release are plotted in the GTM latent space at the location of the mean node responsibility, $\hat{\xi}_n$. Points are colored according to the NDVI computed from the original reflectance spectra. The locations of exemplar spectra for algae, rhodamine, water, and ground points in the latent space are included as color-filled circles. Spectra from the shore and water are clearly separated to different regions of the latent space.\label{fig:gtm-ndvi}}
\end{figure}

@@

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=17.0cm]{paper/figures/results/spectral-signatures.png}
\end{adjustwidth}
\caption{Spectral signatures $\psi(\xi_k)$ corresponding to GTM nodes with non-zero responsibility for exemplar spectra corresponding to the rhodamine dye plume (top left), algae (top right), ground (bottom left), and open water (bottom right). A pseudo-color image is inset into each plot with the location of the exemplar spectrum marked with a white circle. \label{fig:spectral-signatures}}
\end{figure}

@@

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=17.0cm]{paper/figures/results/algae.png}
\end{adjustwidth}
\caption{Using the spectral endmember assigned to algae from the trained GTM together with the NS3, we are able to estimate the algal abundance near the shore. (\textbf{Top}) NS3 values of pixels below a threshold of 0.4275 corresponding to a total area of $587.3$ $m^2$. (\textbf{Bottom}) A picture of the pond near the shore showing the algae.\label{fig:algae-map}}
\end{figure}

Spectral endmembers extracted from the GTM can be used to map the abundance of water constituents in HSI by identifying pixels with a NS3 value below a given threshold. In Fig~\ref{fig:algae-map} we demonstrate this by using the extracted algae endmember spectrum to map the abundance of filamentous blue-green algae near the western shore of the pond. \textcolor{red}{The distribution of NS3 values was multi-modal with the majority of values having an NS3 $>0.4$ corresponding to dissimilar spectra. Consequently, a cutoff threshold of $0.4725$ was chosen to distinguish algae values corresponding to the $20$th-quantile of the NS3 distribution.} The estimated area subsumed by algae is shown in Fig~\ref{fig:algae-map} and calculated to be $587.3$ $m^2$. Similarly, we are able to track the evolution of the rhodamine dye release into the western portion of the pond by computing the NS3 for HSI collected across multiple flights. \textcolor{red}{Here the same cutoff threshold was used corresponding to the $33$rd-quantile of the NS3 distribution.} In the left panel of Fig~\ref{fig:rhodamine-map}, we see that the dye plume initially encompasses an area of roughly $330.7$ $m^2$. A second flight performed 15 minutes later reveals the dye plume to have increased to an area of $1495.6$ $m^2$. The increase in NS3 values between flights reflects the dilution of dye as it diffused into the water.

@@

In this study, we explored the GTM as a unsupervised approach to simultaneously enable visualization of HSI and perform endmember extraction of spectral signatures. First, we showed how the representation of data in the latent space of the GTM given by the mean responsibility can be used to visualize the small-scale structures within inland water bodies as evidenced in Fig~\ref{fig:gtm-water}. In particular, we note that the sharp gradients in colors found when visualizing the spatial distribution of GTM nodes across the pond reflects significant variability in water composition at the sub-meter scale. This has important consequences for water quality assessment where the location of in situ data collection will have a strong impact on the ability of any model to predict water quality parameters from HSI. Visualizing the distribution of spectra from collected HSI is therefore highly relevant to guide in situ data collection as shifting the sampling location by as little as a meter can lead to significant differences in measured values.

@@

The second application of the GTM presented in this study is for the unsupervised extraction of endmembers corresponding to unique sources observed in the HSI. Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenarios \cite{nonlinearity-in-hsi}. Additionally, because the GTM is a probabilistic model, the values of model hyperparameters can be selected objectively using information criteria like the BIC. This is a clear advantage over similar methods like the SOM on which the GTM is based. Endmembers identified for exemplar spectra corresponding to rhodamine, algae, ground, and water pixels demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources. If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform as semi-supervised classification similar to the SOM approach developed by Riese et al. \cite{riese2019supervised}.

@@

The primary limitation of the GTM is that the number of nodes increases exponentially with the dimension of the latent space. However when constrained to two dimensions in order to enable visualization of the resulting map, the number of latent nodes has a negligible impact compared to the size of the dataset on which it is trained. Additionally, the GTM considers individual HSI pixels and does not exploit spatial structure like other methods such as convolutional autoencoders and non-negative matrix factorization using super-pixels \cite{palsson2020convolutional,Feng2022HyperspectralUB}. Extensions to the GTM have been proposed to enable batch training for large datasets as well as manifold-aligned noise models which replace the fixed variance parameter $\beta^{-1}$ with a full covariance matrix \cite{gtm-bishop-2}.

@@

In this study we present the GTM as an useful unsupervised method for the visualization of UAV-based hyperspectral imagery and associated extraction of spectral endmembers. Using data collected at a North Texas pond, we demonstrate how the latent space of the GTM can be used to visualize the distribution of observed reflectance spectra revealing the small-scale spatial variability of water composition. Spectral signatures extracted from GTM nodes are used to successfully map the abundance of algae near the shore and to track the evolution of a rhodamine tracer dye plume. These examples illustrate the power of combining unsupervised learning with UAV-based HSI collection for the characterization of water composition.

>>>>

In this paper, we explore the application of the GTM to UAV-acquired HSI for the characterization water quality. Using data collected at a pond in Montague, North Texas, we first train a GTM using water-only pixels to produce a low-dimensional representation of the collected HSI. We use this mapping to explore the highly detailed small-scale variability within the pond and discuss how this can be used to guide reference data collection. Next, we demonstrate how the GTM can be utilized to identify relevant spectral endmembers from a combined dataset including ground pixels, near-shore filamentous blue-green algae, water, and a simulated contaminant plume using a rhodamine tracer dye. Once identified, these endmembers can be used to rapidly map the abundance of spectral features within the pond. We demonstrate this capability by using a GTM to map the abundance of filamentous blue-green algae near the shore as well as the dispersion of a rhodamine dye plume.

@@

In this study, we explore the use of the GTM as a tool for dimensionality reduction and unsupervised endmember extraction of hyperspectral imagery. To this end, a dataset of HSI was collected at a pond in Montague, North Texas on 23 November 2020 using a UAV mounted hyperspectral imager configured as described in \cite{robot-team-1, robot-team-2}. The pond spans an area $<0.1$ $\text{km}^2$ and has a maximum depth of $3$ m. Across multiple flights, many data cubes like the example shown in panel a of Fig~\ref{fig:sample-spectra} were collected Additionally, rhodamine dye was released and subsequently imaged to simulate a localized contaminant source. Using the HSI shown in Fig~\ref{fig:sample-spectra}, a set of exemplar spectra were visually identified corresponding to near-shore filamentous blue-green algae, water, rhodamine dye, and the ground (a combination of soil and dry grass). The location of these samples is shown in panel b and their corresponding reflectance spectra are visualized in panel c.  In the remainder of this section we first provide a detailed overview of the GTM algorithm. Next we describe the UAV platform used for HSI collection and the various steps employed to process captured HSI. Finally, we describe the two case studies presented in this paper for the utilization of the GTM as a HSI visualization tool and as an endmember extraction technique.

@@

Points $\xi$ in this latent space are mapped to the embedded manifold by a non-linear function $\psi$ parameterized by weights $W$. However, since real data is rarely noise-free, this embedded manifold will not be perfectly thin. To account for this, the points $\xi$ are described in the data space by a radially symmetric Gaussian distribution, $\mathcal{N}(\psi(\xi),\, \beta^{-1})$, with mean $\psi(\xi)$ and precision $\beta$. For a dataset $\left\{ \mathbf{x}_n\right\}\limits_{n=1}^N$ consisting of $N$-many records, this choice yields a log likelihood function of the form
\begin{equation}\label{eq:llh}
    \mathcal{L}(W, \beta) = \sum_n^N \ln \left(\dfrac{1}{K}\sum_k^K p(\mathbf{x}_n \mid \xi_k, W, \beta) \right)
\end{equation}
which can be maximized to obtain optimal $W$ and $\beta^{-1}$.

@@

Slices $R_{[:,n]}$ of the matrix $R$ define the responsibility of each latent node $\xi_k$ for the $n$th data record $\mathbf{x}_n$. Therefore the final responsibility matrix after GTM training can be used to represent each record in the latent space by the mean:
\begin{equation}\label{eq:mean-resp}
    \hat{\xi}_n = \sum_{k}^K R_{kn}\xi_k.
\end{equation}

@@

To account for the variability of incident light, raw HSI are converted to units of reflectance using the downwelling irradiance spectrum simultaneously captured with each HSI. With the hyperspectral imager oriented to nadir, the reflectance is given by
\begin{equation}\label{eq:reflectance}
    \rho(\lambda) = \pi L(\lambda)/E_d(\lambda)
\end{equation}
where $L$ is the spectral radiance, $E_d$ is the downwelling irradiance, and a factor of $\pi$ steradians results from assuming Lambertian (diffuse) upwelling radiance \cite{ruddick2019review}. HSI collection was performed near solar noon to maximize the amount of incident sunlight illuminating the water. For the site in North Texas, this corresponded to an average solar zenith angle of $54.9^\circ$ for 23 November 2020. During flights, the hyperspectral imager was oriented to nadir resulting in HSI with negligible sunglint effects.

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=15.5cm]{paper/figures/methods/annotated-drone.pdf}
\end{adjustwidth}
\caption{The UAV platform: (\textbf{a}) The Resonon Pika XC2 hyperspectral imager. (\textbf{b}) The Freefly Alta-X with the attached hyperspectral imager, processing computer, and downwelling irradiance spectrometer. \label{fig:drone}}
\end{figure}

@@

Next, we consider a dataset of combined HSI pixels including both water and land. To simulate the dispersion of a potential contaminant source, a rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume. From these HSI a collection of $>145,000$ pixels were sampled for model training. To demonstrate the ability to extract relevant spectral endmembers from the trained GTM, we utilize the exemplar spectra corresponding to open water, ground, near-shore algae, and rhodamine dye visually identified in Fig~\ref{fig:sample-spectra}. The responsibility matrix obtained by the trained GTM provides a probability distribution over the latent space nodes for every data record. We identify relevant endmembers for each reference spectrum via the set of nodes with non-zero responsibility. We then utilize the mapping $\psi$ to obtain the spectral representation of these nodes. Since the GTM is unsupervised, these extracted endmembers do not correspond to individual records, but rather spectral features learned by the GTM.

Values for the hyperparameters $m$, $s$, and $\alpha$, are determined by training multiple GTM models and selecting values which minimize the Bayesian Information Criterion (BIC) given by 
\begin{equation}
    \text{BIC} = P\ln(N) - 2\mathcal{L}
\end{equation}
where $P$ is the total number of model parameters, $N$ is the number of records in the dataset, and $\mathcal{L}$ is the log likelihood defined in Eq~\ref{eq:llh}.

The normalized spectral similarity score (NS3) introduced by Nidamanuri and Zbell combines the root-mean-square (RMS) difference together with the spectral angle to provide a spectral distance function\cite{nidamanuri2010normalized}. For two spectra $\rho_1(\lambda)$ and $\rho_2(\lambda)$ it is defined by
\begin{equation}
    \text{NS3}(\rho_1, \rho_2) &= \sqrt{RMS(\rho_1, \rho_2)^2 + (1-\cos\theta)^2}
\end{equation}
where
\begin{align}
    \text{RMS}(\rho_1, \rho_2) &= \sqrt{\frac{1}{N-1}\sum_\lambda \left(\rho_{1}(\lambda) - \rho_2(\lambda) \right)^2} \\
    \cos\theta &= \frac{\langle \rho_1 , \rho_2 \rangle}{\lVert \rho_1\rVert \lVert \rho_2 \rVert}.
\end{align}

We use the NS3 together with extracted GTM endmembers to map the abundance of filamentous blue-green algae near the shore as well as the evolution of the rhodamine dye plume. To do this, the GTM node with maximum responsibility is selected from the endmembers identified for the algae and rhodamine reference spectra. The distribution of NS3 values between each endmember and all HSI pixels is then used to select a suitable cutoff threshold. The area subsumed by HSI pixels with an NS3 below this threshold is then used to infer the spatial extent of algae and rhodamine in the pond.

% add something about thresholding here...

@@

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=18.0cm]{paper/figures/results/gtm-water.png}
\end{adjustwidth}
\caption{Visualization of the GTM trained solely on water pixels (no land and no rhodamine plume). Each data point is colored by computing the mean latent space position, $\hat{\xi}_n$ according to Eq~\ref{eq:mean-resp} with red corresponding to the $\xi_1$ coordinate and blue corresponding to the $\xi_2$ coordinate. \textbf{(a)} Spatial distribution learned by the GTM. \textbf{(b)} The distribution of data in the latent space. Nodes corresponding to the four corners and center of the latent space are identified with cyan markers. \textbf{(c)} reflectance spectra corresponding to the selected GTM nodes computed via the nonlinear mapping $\psi(\xi_k$).\label{fig:gtm-water}}
\end{figure}

@@

To visualize the distribution of HSI learned by the GTM, we can associate a color with each dimension of the latent space. In Fig~\ref{fig:gtm-water} we have used the red channel to represent $\xi_1$ and the blue channel to represent $\xi_2$. Applying the trained GTM to compute mean node responsibilities for all water-pixels in collected HSI allows us to illustrate the spatial distribution of the spectra on a map as shown in panel (a). Here we observe a clear distinction between water near the shore and water in the middle of the pond. Additionally, the eastern alcove of the pond is significantly more blue than the rest of the water. We note that flow through this region is restricted by the small entrance near the cyan square shown in panel a. However, despite its similar depth and surface characteristics, the GTM has distinguished this region from the rest of the water.

The close proximity of highly dissimilar GTM classes illustrated by sharp color gradients in the map reflects the small-scale spatial variability typical of inland water bodies like this pond. For example, water near the cyan diamond in panel a of Fig~\ref{fig:gtm-water} includes both blue and red pixels in close proximity. As shown in panel b, these colors correspond to the top-left and bottom-right corners of the GTM latent space. As the GTM model maps neighboring nodes to similar spectra, this region includes maximally dissimilar pixels. Furthermore, this point is away from the shore where the depth is approximately uniform. The precision parameter $\beta$ learned by the GTM enables the model to account for spectral variability so that effects from localized disturbances such as surface ripples are constrained to small displacements in the latent space. With these considerations in mind, contrast in this region of the pond therefore reflects real variability of water composition, surface sediments, and vegetation.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\columnwidth]{paper/figures/results/square-ndvi.pdf}
\caption{GTM latent space visualization: Each sample spectrum from the dataset of combined HSI covering shore, water, and the rhodamine dye release are plotted in the GTM latent space at the location of the mean node responsibility, $\hat{\xi}_n$. Points are colored according to the NDVI computed from the original reflectance spectra. The locations of exemplar spectra for algae, rhodamine, water, and ground points in the latent space are included as color-filled circles. Spectra from the shore and open water are clearly separated to different regions of the latent space.\label{fig:gtm-ndvi}}
\end{figure}

@@

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=17.0cm]{paper/figures/results/spectral-signatures.png}
\end{adjustwidth}
\caption{Spectral signatures $\psi(\xi_k)$ corresponding to GTM nodes with non-zero responsibility for the rhodamine dye plume (top left), near-shore algae (top right), ground (bottom left), and open water (bottom right). A pseudo-color image is inset into each plot with the location of the exemplar spectrum marked with a white circle. \label{fig:spectral-signatures}}
\end{figure}

@@

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=17.0cm]{paper/figures/results/algae.png}
\end{adjustwidth}
\caption{Using the spectral endmember assigned to algae from the trained GTM together with the NS3, we are able to estimate the algal abundance near the shore. (\textbf{Left}) NS3 values of pixels below a threshold of 0.4275 corresponding to a total area of $587.3$ $m^2$. (\textbf{Right}) A picture of the pond near the shore showing the algae.\label{fig:algae-map}}
\end{figure}

Spectral endmembers extracted from the GTM can be used to map the abundance of water constituents in HSI by identifying pixels with a NS3 value below a given threshold. In Fig~\ref{fig:algae-map} we demonstrate this by using the extracted algae endmember spectrum to map the abundance of filamentous blue-green algae near the western shore of the pond. \textcolor{red}{The distribution of NS3 values was multi-modal with a majority of pixels above $0.4$ belonging to dissimilar spectra. Consequently, a cutoff threshold of $0.4275$ was chosen to distinguish algae values corresponding to the $20$th-quantile of the NS3 distribution.} The estimated area subsumed by algae is shown in Fig~\ref{fig:algae-map} and calculated to be $587.3$ $m^2$. Similarly, we are able to track the evolution of the rhodamine dye release into the western portion of the pond by computing the NS3 for HSI collected across multiple flights. \textcolor{red}{Here the same cutoff threshold was used corresponding to the $33$rd-quantile of the NS3 distribution.} In the left panel of Fig~\ref{fig:rhodamine-map}, we see that the dye plume initially encompasses an area of roughly $330.7$ $m^2$. A second flight performed 15 minutes later reveals the dye plume to have increased to an area of $1495.6$ $m^2$. The increase in NS3 values between flights reflects the dilution of dye as it diffused into the water.

@@

In this study, we explored the GTM as a unsupervised approach to simultaneously enable visualization of HSI and perform endmember extraction of spectral signatures. First, we showed how the representation of data in the latent space of the GTM given by the mean responsibility can be used to visualize the small-scale structures within inland water bodies as evidenced in Fig~\ref{fig:gtm-water}. In particular, we note that the sharp gradients in colors found when visualizing the spatial distribution of GTM nodes across the pond reflects significant spectral variability at the sub-meter scale. This has important consequences for water quality assessment where the location of in situ data collection will have a strong impact on the ability of any model to predict water quality parameters from HSI. Visualizing the distribution of spectra from collected HSI is therefore highly relevant to guide in situ data collection as shifting the sampling location by as little as a meter can lead to significant differences in measured values.

@@

The second application of the GTM presented in this study is for the unsupervised extraction of endmembers corresponding to unique sources observed in the HSI. Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenarios \cite{nonlinearity-in-hsi}. Additionally, because the GTM is a probabilistic model, the values of model hyperparameters can be selected objectively using information criteria like the BIC. This is a clear advantage over similar methods like the SOM on which the GTM is based. Endmembers identified for exemplar spectra corresponding to rhodamine, algae, ground, and water pixels demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources. If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform a semi-supervised classification similar to the SOM approach developed by Riese et al. \cite{riese2019supervised}. For example, in clear waters where variation in sediments and vegetation at the pond floor will contribute to observed spectral variability, the distribution learned by the GTM can be paired with in-situ data to classify the surface type.

@@

The primary limitation of the GTM is that the number of nodes increases exponentially with the dimension of the latent space. However when constrained to two dimensions in order to enable visualization of the resulting map, the number of latent nodes has a negligible impact compared to the size of the dataset on which it is trained. Additionally, the GTM considers individual HSI pixels and does not exploit spatial structure like other methods such as convolutional autoencoders and non-negative matrix factorization using super-pixels \cite{palsson2020convolutional,Feng2022HyperspectralUB}. Extensions to the GTM have been proposed to enable batch training for large datasets as well as manifold-aligned noise models which replace the fixed precision parameter $\beta$ with a full covariance matrix \cite{gtm-bishop-2}.

@@

In this study we present the GTM as a useful unsupervised method for the visualization of UAV-based hyperspectral imagery and associated extraction of spectral endmembers. Using data collected at a North Texas pond, we demonstrate how the latent space of the GTM can be used to visualize the distribution of observed reflectance spectra revealing the small-scale spatial variability of water composition. Spectral signatures extracted from GTM nodes are used to successfully map the abundance of algae near the shore and to track the evolution of a rhodamine tracer dye plume. These examples illustrate the power of combining unsupervised learning with UAV-based HSI collection for the characterization of water composition.
